---
title: "Dynamic Data @ useR"
author: "Jo Hardin - Pomona College"
date: "June 30, 2016"
output:
  slidy_presentation:
    font_adjustment: +10
    footer: "Jo Hardin, jo.hardin@pomona.edu"
---

```{r echo=FALSE,message=FALSE, warning=FALSE}
require(tidyr)
require(plyr)
require(mosaic)
require(readr)
```

## Motivation

- Engage students (make statistics relevant to their experiences)  

- Put "data at the center of the curriculum"  [Gould and Cetinkaya Rundel (2013)]  

- Capitalize on the fantastic R packages which have been developed recently  

- Take advantage of the many interesting data sets which are publicly available   

## What is Dynamic Data?

In contrast to (most) *static* data, *dynamic* data...  

- is compiled / kept up to date by someone else  

- is uploaded directly from the internet   

- isn't particularly clean  

- IS accessible at all levels of statistics  


## Project Goals

- To bring dynamic data sets into the curriculum  

- ... in order to answer interesting statistical questions  

- ... and other scientific questions.  

## Project Goals

- To bring dynamic data sets into the curriculum

- ... in order to answer interesting statistical questions

- ... and other scientific questions.

<div class="centered">
**What is new?  Nothing.**
</div>

## Project Goals

- To bring dynamic data sets into the curriculum

- ... in order to answer interesting statistical questions

- ... and other scientific questions.


<div class="centered">
**What is new?  EVERYTHING.**
</div>


## College Scorecard Data

- US institutions of higher education data on cost, debt, completion rates, and post-graduation earning potential  
- collected by the U.S. Department of Education
- compilation of institutional reporting, federal financial aid reports, and tax information  
- well documented at [https://collegescorecard.ed.gov/data/documentation/](https://collegescorecard.ed.gov/data/documentation/)
- caveat: some of the variables have been collected only on students receiving federal financial aid (biases inherent to analyses done on data collected from such a subgroup should be considered)



## Downloading the College Scorecard Data

```{r cache=TRUE}
college_url <- "https://s3.amazonaws.com/ed-college-choice-public/Most+Recent+Cohorts+(All+Data+Elements).csv"
college_data <- read_csv(college_url)
dim(college_data)
names(college_data)[37:45]
```


## Cleaning the College Scorecard Data

Very big data set, lots of cleaning to do.

```{r cache=TRUE}
college_debt = college_data %>% 
  select(INSTNM,STABBR,PREDDEG, HIGHDEG, region, LOCALE, CCUGPROF,HBCU,
         WOMENONLY, RELAFFIL,ADM_RATE, SATVRMID, SATMTMID,SATWRMID,
         SAT_AVG, UG,NPT4_PUB, NPT4_PRIV, COSTT4_A, DEBT_MDN, CUML_DEBT_P90, 
         mn_earn_wne_p10,md_earn_wne_p10) %>%
  mutate(region2 =  ifelse(region=="0", "Military", 
                    ifelse(region=="1", "New England",
                    ifelse(region=="2", "Mid East", 
                    ifelse(region=="3", "Great Lakes",
                    ifelse(region=="4", "Plains", 
                    ifelse(region=="5", "Southeast",
                    ifelse(region=="6", "Southwest", 
                    ifelse(region=="7", "Rocky Mnts",
                    ifelse(region=="8", "Far West", "Outlying")))))))))) %>%
  mutate(ADM_RATE = extract_numeric(ADM_RATE),
       SATVRMID = extract_numeric(SATVRMID),
       SATMTMID = extract_numeric(SATMTMID),
       SATWRMID = extract_numeric(SATWRMID),
       SAT_AVG = extract_numeric(SAT_AVG),
       UG = extract_numeric(UG),
       NPT4_PUB = extract_numeric(NPT4_PUB),
       NPT4_PRIV = extract_numeric(NPT4_PRIV),
       COSTT4_A = extract_numeric(COSTT4_A),
       DEBT_MDN = extract_numeric(DEBT_MDN),
       CUML_DEBT_P90 = extract_numeric(CUML_DEBT_P90),
       mn_earn_wne_p10 = extract_numeric(mn_earn_wne_p10),
       md_earn_wne_p10 = extract_numeric(md_earn_wne_p10)) %>%
  mutate(RELAFFIL = ifelse(RELAFFIL=="NULL", NA, RELAFFIL),
         LOCALE = ifelse(LOCALE =="NULL", NA, LOCALE),
         CCUGPROF = ifelse(CCUGPROF=="NULL", NA, CCUGPROF),
         HBCU = ifelse(HBCU=="NULL", NA, HBCU),
         WOMENONLY = ifelse(WOMENONLY=="NULL", NA, WOMENONLY))
  

str(college_debt)
summary(college_debt)
```


## Confidence versus Predictions Intervals:  Debt & Income

```{r echo=FALSE}
debt_mod <- lm(DEBT_MDN~1, data = college_debt)
debt_fun <- makeFun(debt_mod)

earn_mod <- lm(md_earn_wne_p10~1, data = college_debt)
earn_fun <- makeFun(earn_mod)
```


```{r echo=FALSE}
#creating the models for building confidence and prediction intervals:
debtreg_mod <- lm(DEBT_MDN~as.factor(region), data = college_debt)
debtreg_fun <- makeFun(debtreg_mod)
earnreg_mod <- lm(md_earn_wne_p10~as.factor(region), data=college_debt)
earnreg_fun <- makeFun(earnreg_mod)

# creating a dataframe for holding the information needed to plot

worth <- data.frame(fit = double(),
                    lowerbound = double(),
                    upperbound = double(),
                    cost = character(),
                    type = character(),
                    regNum = character(),
                    regName = character(),
                    stringsAsFactors = FALSE)

worth[1,] <- c(debt_fun(interval="conf"), "debt", "conf", "all", "US (all)")
worth[2,] <- c(debt_fun(interval="pred"), "debt", "pred", "all", "US (all)")
worth[3,] <- c(earn_fun(interval="conf"), "earn", "conf", "all", "US (all)")
worth[4,] <- c(earn_fun(interval="pred"), "earn", "pred", "all", "US (all)")

for(i in 0:9){
  worth <- rbind(worth, 
                 c(debtreg_fun(region=i,interval="conf"), "debt","conf",
                   i,college_debt[college_debt$region==i,]$region2[1]))
  worth <- rbind(worth, 
                 c(debtreg_fun(region=i,interval="pred"), "debt","pred",
                   i,college_debt[college_debt$region==i,]$region2[1]))

  worth <- rbind(worth, 
                 c(earnreg_fun(region=i,interval="conf"), "earn","conf",
                   i,college_debt[college_debt$region==i,]$region2[1]))
  worth <- rbind(worth, 
                 c(earnreg_fun(region=i,interval="pred"), "earn","pred",
                   i,college_debt[college_debt$region==i,]$region2[1]))
  }

worth <- worth %>% mutate(fit = extract_numeric(fit),
                          lowerbound = extract_numeric(lowerbound),
                          upperbound = extract_numeric(upperbound))
```

```{r echo=FALSE, fig.align="center", fig.width=8, fig.height=5}
pd <- position_dodge(width = 0.5)
ggplot(worth, aes(x=regName, y=fit)) + 
  geom_point(aes(col=cost), position=pd) +
  geom_errorbar(aes(ymin=lowerbound, ymax=upperbound, col=cost,
                    lty=type), position=pd) + 
  xlab("Region") + ylab("$debt or $income 10 years out") +
  theme(text = element_text(size=8))


```

## NHANES Data

The NHANES data (from the National Health and Nutrition Examination Survey - nationwide survey of CDC)

1. Selection of primary sampling units (PSUs) (counties or small groups of contiguous counties)  
2. Selection of segments within PSUs (cluster of households)  
3. Selection of specific households within segments  
4. Selection of individuals within a household  


```{r warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
require(Hmisc)
NHANES.demo <- sasxport.get("http://wwwn.cdc.gov/Nchs/Nhanes/2011-2012/DEMO_G.XPT")
NHANES.body <- sasxport.get("http://wwwn.cdc.gov/Nchs/Nhanes/2011-2012/BMX_G.XPT")
NHANES.demo <-  
  mutate(NHANES.demo, gender = ifelse(NHANES.demo$riagendr==1, "male", "female")) 

NHANES.comb <-  
  inner_join(NHANES.body, NHANES.demo, by = "seqn")

head(NHANES.comb)
```

## (re-)Sampling the NHANES data

Because the NHANES data were collected using a cluster sampling scheme, it is important to use the variables which describe the weights on the sampling to create a sample which is reflective of the population.  


```{r}
numobs = 5000
SRSsample <- sample(1:nrow(NHANES.comb), numobs, replace=FALSE,
       prob=NHANES.comb$wtmec2yr/sum(NHANES.comb$wtmec2yr))
NHANES.comb <- NHANES.comb[SRSsample,]
```

## Downloading the NHANES Data

```{r warning=FALSE, message=FALSE}
require(Hmisc)
NHANES.demo <- sasxport.get("http://wwwn.cdc.gov/Nchs/Nhanes/2011-2012/DEMO_G.XPT")
NHANES.body <- sasxport.get("http://wwwn.cdc.gov/Nchs/Nhanes/2011-2012/BMX_G.XPT")
NHANES.demo <-  
  mutate(NHANES.demo, gender = ifelse(NHANES.demo$riagendr==1, "male", "female")) 

NHANES.comb <-  
  inner_join(NHANES.body, NHANES.demo, by = "seqn")

head(NHANES.comb)
```


## NHANES in Intro Stats

```{r}
Adults = NHANES.comb %>% 
  filter(ridageyr >=18, bmxbmi>1) %>% 
  filter(dmdmartl>0 & dmdmartl < 10) %>% 
  mutate(rel=ifelse(dmdmartl==6|dmdmartl==1, "committed", "not")) %>%
  mutate(bmi=bmxbmi)
  
bwplot(bmi ~ rel, data=Adults, xlab="Relationship Status", ylab="BMI")

ggplot(Adults, aes(rel, bmi))+ geom_violin(color="orange")+ 
  xlab("Relationship Status") + ylab("BMI")

dim(Adults)
t.test(bmi ~ rel, data=Adults)
```

## NHANES beyond Intro Stats

In a linear models or computational statistics class, smooth curves might be a topic of discussion:

```{r}
ggplot(Adults, aes(x=bmxht, y=bmxwt, group=gender, color=gender)) + geom_point(alpha=.5)+ 
  xlab("Height") + ylab("Weight") + ggtitle("Height vs Weight by Gender")
  
ggplot(Adults, aes(x=bmxht, y=bmxwt, group=gender, color=gender)) + 
  xlab("Height") + ylab("Weight") + geom_point(alpha=.5)+ 
  stat_smooth(alpha=1)+ 
  ggtitle("Height vs Weight by Gender with Smooth Regression Fit")
```


## Manuscript

- Format:
    + Example (description, documentation, references)
    + Using dynamic data in an introductory classroom
    + Thinking outside the box
    
- Full R Markdown files for downloading: https://github.com/hardin47/DynamicData
- Manuscript on arXiv: http://arxiv.org/abs/1603.04912


## Full Examples

- College Scorecard  
- NHANES (National Health and Nutrition Examination Survey)  
- Wikipedia (scraping XML / HTML)  
- Gapminder (Literacy Rates, using **tidyr**'s **gather** function)  
- NOAA weather data from buoys  

## Additional Examples

- Baseball Data  (Jim Albert)  
- Cherry Blossom Ten Mile Run (Kaplan, in Nolan & Temple Lang, 2015)  
- Fatal Accidents & US Census Data (Laura Kapitula)  
- Climate Data  
- Iowa Liquor Sales  
- Medicare Inpatient Charges  



## Thank you!

Jo Hardin  
Pomona College  
jo.hardin@pomona.edu  
https://github.com/hardin47/DynamicData  
@jo_hardin47  

